---
title: "Evaluating quantitative models with modelskill"
author: "modelskill contributors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Evaluating quantitative models with modelskill}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 6)
```

## Why use several diagnostics?

A single statistic rarely describes every important aspect of a quantitative
model. A model may have a strong correlation but still be biased, or it may
have a small average error while failing to reproduce the observed variation.
`modelskill` combines several complementary indices and diagrams so that these
behaviours can be inspected together.

The approach follows Wadoux, Walvoort, and Brus (2022), who introduced the use
of Taylor and solar diagrams for integrated evaluation of quantitative maps.
The same tools are useful more generally for quantitative prediction models.

## Reproducible example

Here we create a small observation vector and three predictions: a perfect
prediction, a biased prediction, and a noisy prediction.

```{r data}
library(modelskill)

set.seed(123)
obs <- seq(0, 10, length.out = 100) + rnorm(100, sd = 1)
mods <- list(
  Good = obs + rnorm(100, sd = 0.5),
  Biased = obs + 1,
  Noisy = obs + rnorm(100, sd = 2)
)
```

## Indices

```{r metrics}
metrics <- model_metrics(mods, obs)
metrics
```

ME describes average signed error, while MAE and RMSE summarize error size.
Correlation describes association, NSE compares performance with a mean-value
benchmark, and `rhoC` combines correlation with location and scale agreement.

## Taylor diagram

The Taylor diagram displays correlation through the polar angle and the
standard-deviation ratio through the radial distance. The reference point is
the red point at correlation 1 and standardized standard deviation 1.

```{r taylor}
gg_taylor(mods, obs, label = TRUE)
```

## Solar diagram

The solar diagram combines normalized mean error and standardized unbiased
RMSD. The point colour can display any model-level statistic, such as NSE.

```{r solar}
gg_solar(mods, obs, colorval = metrics$NSE, colorval.name = "NSE",
         label = TRUE, x.axis_begin = -1.5, x.axis_end = 1.5,
         y.axis_end = 2, by = 0.25)
```

## Target diagram

The target diagram separates signed error components while retaining RMSD
contours. Models closer to the origin generally have smaller combined error,
with the sign of the horizontal coordinate indicating the direction of the
standard-deviation difference.

```{r target}
gg_target(mods, obs, colorval = metrics$NSE, colorval.name = "NSE",
          label = TRUE, axis_begin = -2, axis_end = 2, by = 0.5)
```

## Standard ggplot2 customisation

Each diagram is an ordinary `ggplot2` object, so standard layers can be added.

```{r customise}
gg_taylor(mods, obs) +
  ggplot2::labs(title = "Model comparison") +
  ggplot2::theme_minimal()
```

## References

Wadoux, A. M. J.-C., Walvoort, D. J. J., and Brus, D. J. (2022). An
integrated approach for the evaluation of quantitative soil maps through
Taylor and solar diagrams. *Geoderma*, 405, 115332.
<https://doi.org/10.1016/j.geoderma.2021.115332>
