---
title: "Evaluating quantitative models with modelskill"
author: "modelskill contributors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Evaluating quantitative models with modelskill}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 6)
```

## Why use several diagnostics?

A single statistic rarely describes every important aspect of a quantitative
model. A model may have a strong correlation but still be biased, or it may
have a small average error while failing to reproduce the observed variation.
`modelskill` combines point-prediction statistics, predictive-uncertainty
diagnostics, and graphical summaries so that these behaviours can be inspected
together.

The approach follows Wadoux, Walvoort, and Brus (2022), who introduced the use
of Taylor and solar diagrams for integrated evaluation of quantitative maps.
The same tools are useful more generally for quantitative prediction models.

## Reproducible example

Here we create a small, reproducible reference series and three deliberately
different prediction models. The models are simple on purpose: they make the
meaning of the diagrams easy to see.

```{r data}
library(modelskill)

set.seed(123)
obs <- seq(0, 10, length.out = 100) + rnorm(100, sd = 1)
mods <- list(
  # Follows the observations closely, with only small random errors.
  Good = obs + rnorm(100, sd = 0.5),
  # Has exactly the same pattern, but is systematically one unit too high.
  Biased = obs + 1,
  # Has larger random errors and therefore reproduces the pattern less well.
  Noisy = obs + rnorm(100, sd = 2)
)
```

In a real application, `obs` would be measured values at validation
locations, and each element of `mods` would be predictions at those same
locations. The `Good` model should be closest to the ideal reference. The
`Biased` model illustrates why correlation alone is insufficient: it can have
correlation near one while still having a substantial mean error. The `Noisy`
model illustrates loss of pattern agreement and larger prediction errors.

## Indices

```{r metrics}
metrics <- model_metrics(mods, obs)
metrics
```

The columns have complementary meanings:

* `ME` is the signed average error; its sign indicates systematic
  overprediction or underprediction.
* `MAE` and `RMSE` measure error magnitude, with RMSE giving more weight to
  large errors.
* `r` measures agreement in pattern or ordering, not agreement in level.
* `NSE` compares the predictions with a mean-value benchmark; values closer to
  one indicate better performance relative to that benchmark.
* `rhoC` combines correlation with location and scale agreement.

The diagrams show these complementary properties together rather than asking
the reader to rank models from one statistic alone.

The same calculations are available individually when a single, clearly named
statistic is needed.

```{r individual-metric}
rmse(obs, mods$Good)
ccc(obs, mods$Good)
```

## Predictive uncertainty

Good point predictions do not automatically imply good uncertainty estimates.
Uncertainty validation should assess **calibration** (whether stated intervals
contain observations as often as intended) and **sharpness** (whether intervals
are usefully narrow). Here we create simple 80% and 95% intervals around the
`Good` predictions for illustration.

```{r uncertainty-data}
predictive_sd <- rep(0.8, length(obs))
lower80 <- mods$Good - 1.282 * predictive_sd
upper80 <- mods$Good + 1.282 * predictive_sd
lower95 <- mods$Good - 1.96 * predictive_sd
upper95 <- mods$Good + 1.96 * predictive_sd
```

Interval validation returns prediction interval coverage probability (PICP),
its departure from nominal coverage, average width, and the proper interval
score (lower is better). PICP is the proportion of test observations inside an
interval; for a 95% interval, a calibrated model has PICP close to 0.95.

```{r interval-metrics}
uncertainty_metrics(obs, lower = lower95, upper = upper95, level = .95)
```

For complete predictive distributions, use QCP and PIT to assess calibration,
and CRPS to compare competing probabilistic models. Lower CRPS is better.

```{r distribution-metrics}
qhat <- cbind(`0.1` = mods$Good - 1.282 * predictive_sd,
              `0.5` = mods$Good,
              `0.9` = mods$Good + 1.282 * predictive_sd)
qcp(obs, qhat, c(.1, .5, .9))
crps(obs, pred = mods$Good, predictive_sd = predictive_sd)
```

The calibration curve compares PICP at multiple nominal levels. It is a PICP
**reliability plot**, also called an **accuracy plot** in geostatistical
uncertainty validation (Goovaerts, 2001): it extends a single PICP across the
predictive distribution. When predictive means and predictive standard
deviations are available, `gg_coverage()` can calculate every 1%--99% normal
interval directly. Central PICP does not by itself reveal asymmetric lower- and
upper-tail misses, so it should be complemented by residual and tail-specific
checks when a one-sided bias is relevant.

```{r coverage}
gg_coverage(
  obs, pred = mods$Good, predictive_sd = predictive_sd
)

accuracy_plot_metrics(obs, pred = mods$Good, predictive_sd = predictive_sd)
```

```{r qcp-pit}
gg_qcp(obs, qhat, c(.1, .5, .9))
gg_pit(pit(stats::pnorm(obs, mean = mods$Good, sd = predictive_sd)))
```

## Taylor diagram

The Taylor diagram displays correlation through the polar angle and the
standard-deviation ratio through the radial distance. The reference point is
the red point at correlation 1 and standardized standard deviation 1.

```{r taylor}
gg_taylor(mods, obs, label = TRUE)
```

## Solar diagram

The solar diagram combines normalized mean error and standardized unbiased
RMSD. The point colour can display any model-level statistic, such as NSE.

```{r solar}
gg_solar(mods, obs, colorval = metrics$NSE, colorval.name = "NSE",
         label = TRUE, x.axis_begin = -1.5, x.axis_end = 1.5,
         y.axis_end = 2, by = 0.25)
```

## Target diagram

The target diagram separates signed error components while retaining RMSD
contours. Models closer to the origin generally have smaller combined error,
with the sign of the horizontal coordinate indicating the direction of the
standard-deviation difference.

```{r target}
gg_target(mods, obs, colorval = metrics$NSE, colorval.name = "NSE",
          label = TRUE, axis_begin = -2, axis_end = 2, by = 0.5)
```

## Standard ggplot2 customisation

Each diagram is an ordinary `ggplot2` object, so standard layers can be added.

```{r customise}
gg_taylor(mods, obs) +
  ggplot2::labs(title = "Model comparison") +
  ggplot2::theme_minimal()
```

## References

Wadoux, A. M. J.-C., Walvoort, D. J. J., and Brus, D. J. (2022). An
integrated approach for the evaluation of quantitative soil maps through
Taylor and solar diagrams. *Geoderma*, 405, 115332.
<https://doi.org/10.1016/j.geoderma.2021.115332>
