---
title: "Evaluating continuous prediction models with modelskill"
author: "Alexandre M.J.-C. Wadoux"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Evaluating continuous prediction models with modelskill}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 6)
```

## Why use several diagnostics?

A single statistic rarely describes every important aspect of a continuous
prediction model. A model can have a strong correlation yet be biased, or have
a small average error while failing to reproduce the observed variation.
`modelskill` combines point-prediction statistics, predictive-uncertainty
diagnostics, and graphical summaries so that these behaviours can be inspected
together.

The package evaluates paired observed and predicted numeric values, regardless
of how predictions were produced. It is suitable for machine-learning
regression models (such as random forests, boosting, neural networks,
support-vector machines, and ensembles), as well as linear, geostatistical,
physical, and process-based models. Maps are one useful application, but are
not a requirement. For an honest estimate of predictive performance, use an
independent validation set or an appropriate resampling scheme.

## Reproducible example

Here we create a small, reproducible reference series and three deliberately
different prediction models. The models are simple on purpose: they make the
meaning of the diagrams easy to see.

```{r data}
library(modelskill)

set.seed(123)
obs <- seq(0, 10, length.out = 100) + rnorm(100, sd = 1)
mods <- list(
  # Follows the observations closely, with only small random errors.
  Good = obs + rnorm(100, sd = 0.5),
  # Has exactly the same pattern, but is systematically one unit too high.
  Biased = obs + 1,
  # Has larger random errors and therefore reproduces the pattern less well.
  Noisy = obs + rnorm(100, sd = 2)
)
```

In a real application, `obs` would be measured values at validation
locations, and each element of `mods` would be predictions at those same
locations. The `Good` model should be closest to the ideal reference. The
`Biased` model illustrates why correlation alone is insufficient: it can have
correlation near one while still having a substantial mean error. The `Noisy`
model illustrates loss of pattern agreement and larger prediction errors.

```{r observed-predicted, fig.height=4.5}
plot_data <- data.frame(
  observed = rep(obs, times = length(mods)),
  predicted = unlist(mods, use.names = FALSE),
  model = rep(names(mods), each = length(obs))
)

ggplot2::ggplot(plot_data, ggplot2::aes(observed, predicted, colour = model)) +
  ggplot2::geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  ggplot2::geom_point(alpha = 0.7, size = 1.5) +
  ggplot2::coord_equal() +
  ggplot2::theme_classic() +
  ggplot2::labs(title = "Observed versus predicted values",
                x = "Observed", y = "Predicted", colour = "Model")
```

The dashed equality line represents perfect predictions. `Good` stays close
to it, `Biased` is systematically above it, and `Noisy` is more dispersed.
This first diagnostic is useful for any continuous predictive model; the
following sections quantify and summarize what it shows.

# Part 1: Point-prediction evaluation

This part introduces complementary statistics for evaluating predicted values:
mean error for systematic error, absolute and squared error measures for error
magnitude, association and concordance measures, and benchmark-relative model
efficiency. No single statistic represents every aspect of predictive quality.

```{r metrics}
metrics <- model_metrics(mods, obs)
metrics
```

The package distinguishes two commonly confused quantities. `nse()`, `mec()`,
and uppercase `R2()` are the same model-efficiency coefficient: they compare
squared prediction error with predicting the observed mean. Lowercase `r2()`
is instead squared Pearson correlation, and measures association rather than
agreement. They should not be interchanged.

For specialised workflows, the extended metrics add robust, scale-normalised,
percentage, and quantile-loss summaries:

```{r extended-metrics}
model_metrics(mods, obs, extended = TRUE)
```

The columns have complementary meanings:

* `ME` is the signed average error; its sign indicates systematic
  overprediction or underprediction.
* `MAE` and `RMSE` measure error magnitude, with RMSE giving more weight to
  large errors.
* `r` measures agreement in pattern or ordering, not agreement in level.
* `NSE` compares the predictions with a mean-value benchmark; values closer to
  one indicate better performance relative to that benchmark.
* `rhoC` combines correlation with location and scale agreement.

The diagrams show these complementary properties together rather than asking
the reader to rank models from one statistic alone.

The same calculations are available individually when a single, clearly named
statistic is needed.

```{r individual-metric}
rmse(obs, mods$Good)
ccc(obs, mods$Good)
```

# Part 2: Evaluation of predictive uncertainty

This part evaluates *quantified predictive uncertainty*, rather than only the
accuracy of point predictions. A prediction interval or distribution makes a
probabilistic statement for each case. For example, a central 90% prediction
interval has lower and upper predictive quantiles at 0.05 and 0.95, and should
contain the future observation in about 90% of comparable validation cases.

Good point predictions do not automatically imply good uncertainty estimates.
Schmidinger and Heuvelink (2023) distinguish two complementary properties:

* **Calibration** (also called reliability) asks whether stated probabilities
  agree with observed frequencies.
* **Sharpness** asks whether predictive distributions are concentrated or,
  equivalently for an interval, usefully narrow. Sharpness is meaningful only
  when considered together with calibration: a narrow interval is not useful
  if it misses observations too often.

These principles apply to uncertainty estimates from machine-learning models,
such as quantile regression, ensembles, and conformal intervals, as well as
Bayesian, geostatistical, and process-based models. The appropriate function
depends on the output available: interval bounds, predicted quantiles, CDF
values, predictive samples, or a normal predictive mean and standard
deviation.

Here we create simple central 80% and 95% normal intervals around the `Good`
predictions solely to illustrate the input format.

```{r uncertainty-data}
predictive_sd <- rep(0.8, length(obs))
lower80 <- mods$Good - 1.282 * predictive_sd
upper80 <- mods$Good + 1.282 * predictive_sd
lower95 <- mods$Good - 1.96 * predictive_sd
upper95 <- mods$Good + 1.96 * predictive_sd
```

## Prediction intervals: calibration, sharpness, and interval score

At one nominal interval level, `uncertainty_metrics()` returns four
complementary quantities:

* **PICP** is the prediction interval coverage probability, bounded from 0 to
  1. Its ideal value is the nominal interval level (for example, 0.95 for a
  95% interval).
* **PICP error** is PICP minus the nominal level, bounded from -1 to 1 with
  ideal value 0. Negative values indicate under-coverage (intervals too narrow
  or over-confident); positive values indicate over-coverage (intervals too
  wide or over-cautious).
* **PIW** is the average prediction-interval width. It has the response unit
  and no universal acceptable threshold: smaller values mean sharper intervals,
  provided PICP is close to the nominal level.
* **Interval score** combines width and the distance of observations outside
  the interval. It is a proper scoring rule with lower values preferred and an
  ideal value of 0. Compare scores only for the same response, validation data,
  and nominal level.

PICP is the proportion of test observations inside an interval; for a 95%
interval, a calibrated model has PICP close to 0.95.

```{r interval-metrics}
uncertainty_metrics(obs, lower = lower95, upper = upper95, level = .95)
```

## Calibration across a distribution

One PICP does not establish calibration of the entire predictive distribution.
`gg_coverage()` evaluates PICP over many nominal levels: a calibration curve
near the 1:1 line is desirable. It is often called a **reliability plot** and
was introduced in geostatistical uncertainty validation as an **accuracy plot**
(Goovaerts, 2001). Its summary areas from `accuracy_plot_metrics()` have ideal
value 0; smaller areas mean a curve closer to the 1:1 line.

PICP alone can hide one-sided bias. For example, a 90% interval may still
cover 90% of observations even when most misses occur below its lower bound.
Use QCP and PIT to inspect this more closely:

* **QCP** at quantile probability \(p\) is between 0 and 1 and is ideal when
  it equals \(p\). Values above \(p\) indicate that the predicted quantile is
  generally too high; values below \(p\) indicate that it is generally too
  low.
* **PIT** values are individually between 0 and 1; there is no preferred value
  for one case. Across many validation cases their histogram should be roughly
  uniform. Systematic departures flag distributional bias or incorrect spread.

For a normal predictive distribution, the following examples calculate QCP,
PIT, and CRPS. CRPS is a proper scoring rule with response units: 0 is ideal,
lower values are better, and values should be compared only on the same
response scale and validation data.

```{r distribution-metrics}
qhat <- cbind(`0.1` = mods$Good - 1.282 * predictive_sd,
              `0.5` = mods$Good,
              `0.9` = mods$Good + 1.282 * predictive_sd)
qcp(obs, qhat, c(.1, .5, .9))
crps(obs, pred = mods$Good, predictive_sd = predictive_sd)
```

When predictive means and predictive standard deviations are available,
`gg_coverage()` can calculate every 1%--99% central normal interval directly.

```{r coverage}
gg_coverage(
  obs, pred = mods$Good, predictive_sd = predictive_sd
)

accuracy_plot_metrics(obs, pred = mods$Good, predictive_sd = predictive_sd)
```

```{r qcp-pit}
gg_qcp(obs, qhat, c(.1, .5, .9))
gg_pit(pit(stats::pnorm(obs, mean = mods$Good, sd = predictive_sd)))
```

## Full-distribution scoring rules

For competing full predictive distributions, scoring rules jointly reward
calibration and sharpness. `crps()` and `median_crps()` accept either equally
weighted predictive samples or a normal mean and predictive standard deviation.
`crps_decomposition()` additionally reports RELI, the non-negative reliability
component of ensemble CRPS, whose ideal value is 0. `log_score()` evaluates
the negative log predictive density at the observations; lower values are
better, but it has no universal numerical range and strongly penalizes events
assigned very low density. Use these scores to rank models evaluated on the
same cases rather than as stand-alone pass/fail thresholds.

# Part 3: Summary diagrams and diagnostic plots

The diagrams below combine complementary statistics into compact visual
diagnostics. They should be read alongside the numerical metrics, not as a
replacement for scientific judgement about the intended use of a model.

## Taylor diagram

The Taylor diagram combines correlation, relative spread, and centred error.
Its radial coordinate is the standard-deviation ratio
\(\sigma^* = \mathrm{sd}(pred) / \mathrm{sd}(obs)\); its polar angle is
\(\arccos(r)\). The distance from a model point to the reference point
\((\sigma^*=1, r=1)\) is the standardized centred root mean square difference:

\[
\mathrm{SDE}^* = \sqrt{1 + \sigma^{*2} - 2\sigma^*r}.
\]

The `Good` point should be near the reference point. `Biased` can also be near
it because a Taylor diagram does **not** represent mean error; its perfect
linear association and unchanged spread hide its constant offset. `Noisy`
should be farther away and at a larger polar angle because it has weaker
pattern agreement. A radius below one indicates smoothing (too little
variation); a radius above one indicates too much variation.

```{r taylor}
gg_taylor(mods, obs, label = TRUE)
```

## Solar diagram

The solar diagram makes the mean error omitted by a Taylor diagram explicit.
It uses the standardized decomposition

\[
\mathrm{RMSE}^{*2} = \mathrm{ME}^{*2} + \mathrm{SDE}^{*2}.
\]

Horizontal position is ME* and vertical position is SDE*. Distance from the
origin is RMSE*. Under `modelskill`'s `obs - pred` convention, points to the
left overpredict and points to the right underpredict. Thus `Biased` falls to
the left, `Good` should be near the origin, and `Noisy` should lie higher on
the SDE* axis. The outer circle at RMSE* = 1 marks the mean-prediction
benchmark: points inside it have positive MEC/NSE/R2. The pale regions indicate
correlation lower bounds, not exact correlation values.

```{r solar}
gg_solar(mods, obs, label = TRUE,
         x.axis_begin = -1.5, x.axis_end = 1.5,
         y.axis_end = 2, by = 0.25)
```

## Target diagram

The target diagram builds on the same RMSE* decomposition but additionally
shows whether prediction variability is lower or higher than observed
variability. The signed SDE coordinate distinguishes smoother from more
variable predictions, while ME* distinguishes overprediction from
underprediction. The distance from the origin remains RMSE*, so `Good` should
be nearest the origin. `Biased` is chiefly displaced by mean error, whereas a
model displaced in the signed-SDE direction chiefly has a spread or pattern
difference. The axis-title placement retains the original diagram style; the
function help documents the underlying coordinates explicitly.

```{r target}
gg_target(mods, obs, label = TRUE,
          axis_begin = -2, axis_end = 2, by = 0.5)
```

## Standard ggplot2 customisation

Each diagram is an ordinary `ggplot2` object, so standard layers can be added.

```{r customise}
gg_taylor(mods, obs) +
  ggplot2::labs(title = "Model comparison") +
  ggplot2::theme_minimal()
```

## References

Wadoux, A. M. J.-C., Walvoort, D. J. J., and Brus, D. J. (2022). An
integrated approach for the evaluation of quantitative soil maps through
Taylor and solar diagrams. *Geoderma*, 405, 115332.
<https://doi.org/10.1016/j.geoderma.2021.115332>

Taylor, K. E. (2001). Summarizing multiple aspects of model performance in a
single diagram. *Journal of Geophysical Research*, 106, 7183-7192.
<https://doi.org/10.1029/2000JD900719>

Jolliff, J. K., Kindle, J. C., Shulman, I., Penta, B., Friedrichs, M. A. M.,
Helber, R., and Arnone, R. A. (2009). Summary diagrams for coupled
hydrodynamic-ecosystem model skill assessment. *Journal of Marine Systems*,
76, 64-82. <https://doi.org/10.1016/j.jmarsys.2008.05.014>

Schmidinger, J. and Heuvelink, G. B. M. (2023). Validation of uncertainty
predictions in digital soil mapping. *Geoderma*, 437, 116585.
<https://doi.org/10.1016/j.geoderma.2023.116585>
