---
title: "Prediction performance metrics"
author: "Alexandre M.J.-C. Wadoux"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Prediction performance metrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
```

Continuous predictions can come from machine learning, statistical models, or
physical models. `modelskill` evaluates the paired predictions and observations,
not the algorithm that produced them. Use an independent validation set or an
appropriate resampling scheme.

```{r data}
library(modelskill)
set.seed(123)
obs <- seq(0, 10, length.out = 100) + rnorm(100, sd = 1)
models <- list(
  Good = obs + rnorm(100, sd = 0.5),
  Biased = obs + 1,
  Noisy = obs + rnorm(100, sd = 2)
)
```

## Start with observed versus predicted values

```{r scatter}
plot_data <- data.frame(
  observed = rep(obs, times = length(models)),
  predicted = unlist(models, use.names = FALSE),
  model = rep(names(models), each = length(obs))
)
ggplot2::ggplot(plot_data, ggplot2::aes(observed, predicted, colour = model)) +
  ggplot2::geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  ggplot2::geom_point(alpha = 0.7, size = 1.5) +
  ggplot2::coord_equal() + ggplot2::theme_classic() +
  ggplot2::labs(x = "Observed", y = "Predicted", colour = "Model")
```

The equality line represents perfect predictions. The `Good` model stays close
to it, `Biased` is shifted upward, and `Noisy` is more dispersed.

## Categories of prediction performance metrics

Each category answers a different question. The table is a practical starting
point for choosing a complementary set of metrics rather than reducing model
performance to a single number.

| Category | `modelskill` metrics | Main question |
|:--|:--|:--|
| **Error magnitude** | RMSE, MAE, MSE, MdAE | How large are the errors? |
| **Bias** | ME / `bias()`, MPE | Are predictions systematically high or low? |
| **Association** | `r`, `r2()` | Do predictions track variation in observations? |
| **Agreement** | Willmott `d`, CCC | How closely do predictions agree in correlation, location, and scale? |
| **Efficiency / skill** | NSE / MEC / `R2()`, KGE | How good are predictions relative to a reference? |
| **Relative / scaled error** | NRMSE, RRMSE, RPD, RPIQ, RER | How large is error relative to the data scale? |

MPE and RRMSE are expressed as percentages. Both are undefined when the
observed value or observed mean used as their denominator is zero, respectively.

## Calculate complementary metrics

```{r metrics}
model_metrics(models, obs)
```

No single metric is sufficient. Mean error (ME) describes signed systematic
error under the package convention `obs - pred`; negative ME indicates
overprediction. MAE and RMSE describe error magnitude in response units; RMSE
gives greater weight to large errors. Pearson correlation describes pattern
association, not agreement in level.

`nse()`, `mec()`, and uppercase `R2()` are aliases for the same
mean-benchmark efficiency coefficient. Lowercase `r2()` is squared Pearson
correlation and must not be interpreted as the efficiency coefficient.
Concordance (`ccc()`) combines correlation with level and spread agreement.

For specialised workflows, request robust, relative, percentage, agreement,
and KGE metrics explicitly:

```{r extended}
model_metrics(models, obs, extended = TRUE)
```

## Equations, interpretation, and references

For the retained pairs, let $o_i$ be observations, $p_i$ predictions,
$e_i=o_i-p_i$, $b=\overline e$, and $n$ the number of pairs.
$\bar o$ and $\bar p$ are the observation and prediction means.
The sample standard deviations are
$s_o=\sqrt{\frac{1}{n-1}\sum_{i=1}^{n}(o_i-\bar o)^2}$ and
$s_p=\sqrt{\frac{1}{n-1}\sum_{i=1}^{n}(p_i-\bar p)^2}$.
For concordance, the population moments are
$v_o=\frac{1}{n}\sum_{i=1}^{n}(o_i-\bar o)^2$,
$v_p=\frac{1}{n}\sum_{i=1}^{n}(p_i-\bar p)^2$, and
$c_{op}=\frac{1}{n}\sum_{i=1}^{n}(o_i-\bar o)(p_i-\bar p)$.
Write $A=\sum_i\lvert o_i-\bar o\rvert$ and
$D=\sum_i(\lvert p_i-\bar o\rvert+\lvert o_i-\bar o\rvert)^2$.
All sums below run over retained pairs. Undefined cases return `NA`.
The listed ranges apply when the metric is defined; there are no universal
cutoffs for acceptable accuracy across applications.

### Core output

The statistic name below matches the left-hand side of its equation; the
output column name is shown in parentheses. Pearson's $r$ is returned in
the `correlation` column.

| Statistic (output column) | Equation | Range, ideal, and interpretation | Reference |
|:--|:--|:--|:--|
| ME (`bias`) | $\mathrm{ME}=\frac{1}{n}\sum_{i=1}^{n}(o_i-p_i)$ | Real; 0 ideal. Positive: underprediction; errors can cancel. Response units. | Legates & McCabe (1999) |
| MAE (`mae`) | $\mathrm{MAE}=\frac{1}{n}\sum_{i=1}^{n}\lvert o_i-p_i\rvert$ | Nonnegative; 0 ideal. Average absolute error, response units. | Willmott & Matsuura (2005) |
| MSE (`mse`) | $\mathrm{MSE}=\frac{1}{n}\sum_{i=1}^{n}(o_i-p_i)^2$ | Nonnegative; 0 ideal. Squared response units. | Hodson (2022) |
| RMSE (`rmse`) | $\mathrm{RMSE}=\sqrt{\frac{1}{n}\sum_{i=1}^{n}(o_i-p_i)^2}$ | Nonnegative; 0 ideal. Response units; sensitive to large errors. | Hodson (2022) |
| NRMSE (`nrmse`) | $\mathrm{NRMSE}=\frac{\mathrm{RMSE}}{s_o}$ | Nonnegative; 0 ideal. 1 means RMSE equals sample observation SD; NA for constant observations. | Taylor (2001); package normalization |
| cRMSE (`crmse`) | $\mathrm{cRMSE}=\sqrt{\frac{1}{n}\sum_{i=1}^{n}[(o_i-p_i)-\mathrm{ME}]^2}$ | Nonnegative; 0 ideal. Removes constant bias; response units. | Taylor (2001) |
| Pearson's $r$ (`correlation`) | $r=\frac{\sum_{i=1}^{n}(o_i-\bar o)(p_i-\bar p)}{\sqrt{\sum_{i=1}^{n}(o_i-\bar o)^2\sum_{i=1}^{n}(p_i-\bar p)^2}}$ | [-1,1]; 1 indicates increasing linear association, not necessarily agreement. NA for either constant vector. | Legates & McCabe (1999) |
| Squared Pearson's $r$, $r^2$ (`r2`) | $r^2=\frac{[\sum_{i=1}^{n}(o_i-\bar o)(p_i-\bar p)]^2}{\sum_{i=1}^{n}(o_i-\bar o)^2\sum_{i=1}^{n}(p_i-\bar p)^2}$ | [0,1]; squared association, loses the sign of correlation. | Legates & McCabe (1999) |
| $R^2$ / NSE / MEC (`R2`) | $R^2=1-\frac{\sum_{i=1}^{n}(o_i-p_i)^2}{\sum_{i=1}^{n}(o_i-\bar o)^2}$ | At most 1; 1 ideal, 0 observed-mean benchmark, negative worse. NA for constant observations. | Nash & Sutcliffe (1970); Janssen & Heuberger (1995) |
| SD ratio (`sd_ratio`) | $\mathrm{SD\ ratio}=\frac{s_p}{s_o}$ | Nonnegative; 1 equal spread, below 1 smoother, above 1 more variable. | Taylor (2001) |
| CCC (`ccc`) | $\mathrm{CCC}=\frac{2c_{op}}{v_o+v_p+(\bar o-\bar p)^2}$ | [-1,1]; 1 ideal agreement. Package returns NA for constant observations and 0 for constant predictions with varying observations. | Lin (1989) |
| $C_b$ (`Cb`) | $C_b=\frac{2\sqrt{v_ov_p}}{v_o+v_p+(\bar o-\bar p)^2}$ | [0,1]; 1 equal means and SDs, irrespective of correlation. Same constant-input policy as CCC. | Lin (1989) |

### Extended output (`extended = TRUE`)

Here $Q_{0.25}(o)$ and $Q_{0.75}(o)$ are the observation quartiles, so
$\mathrm{IQR}(o)=Q_{0.75}(o)-Q_{0.25}(o)$. The logarithm is natural.
ME and RMSE are defined in the core table above, and $r$ is Pearson's $r$.

| Statistic (output column) | Equation | Range, ideal, and interpretation | Reference / convention |
|:--|:--|:--|:--|
| MdAE (`mdae`) | $\mathrm{MdAE}=\mathrm{median}_{i=1,\ldots,n}(\lvert o_i-p_i\rvert)$ | Nonnegative; 0 ideal. Robust typical error, response units. | Hyndman & Koehler (2006) |
| RPD (`rpd`) | $\mathrm{RPD}=\frac{s_o}{\mathrm{RMSE}}$ | Nonnegative; larger is better. NA at perfect prediction. | Bellon-Maurel et al. (2010) |
| RPIQ (`rpiq`) | $\mathrm{RPIQ}=\frac{Q_{0.75}(o)-Q_{0.25}(o)}{\mathrm{RMSE}}$ | Nonnegative; larger is better. Type-7 quartiles; NA when RMSE=0. | Bellon-Maurel et al. (2010) |
| SEP (`sep`) | $\mathrm{SEP}=\sqrt{\frac{1}{n-1}\sum_{i=1}^{n}[(o_i-p_i)-\mathrm{ME}]^2}$ | Nonnegative; 0 ideal. Bias removed; at least two pairs. | Bellon-Maurel et al. (2010) |
| RER (`rer`) | $\mathrm{RER}=\frac{\max_i o_i-\min_i o_i}{\mathrm{RMSE}}$ | Nonnegative; larger is better; sensitive to extremes; NA when RMSE=0. | Bellon-Maurel et al. (2010) |
| MAPE (`mape`) | $\mathrm{MAPE}=\frac{1}{n}\sum_{i=1}^{n}\left\lvert\frac{o_i-p_i}{o_i}\right\rvert$ | Nonnegative fraction; 0 ideal. Multiply by 100 for percent; NA for any zero observation. | Hyndman & Koehler (2006) |
| MPE (`mpe`) | $\mathrm{MPE}=\frac{100}{n}\sum_{i=1}^{n}\frac{o_i-p_i}{o_i}$ | Unbounded percent; 0 ideal. Positive means underprediction only for positive observations; NA at zero. | Hyndman & Koehler (2006); obs-minus-pred sign |
| sMAPE (`smape`) | $\mathrm{sMAPE}=\frac{1}{n}\sum_{i=1}^{n}\frac{2\lvert o_i-p_i\rvert}{\lvert o_i\rvert+\lvert p_i\rvert}$ | [0,2] fraction; 0 ideal. Both-zero pair contributes 0; multiply by 100 for percent. | Hyndman & Koehler (2006); absolute-denominator variant |
| MSLE (`msle`) | $\mathrm{MSLE}=\frac{1}{n}\sum_{i=1}^{n}[\log(1+o_i)-\log(1+p_i)]^2$ | Nonnegative; 0 ideal. NA for negative inputs; target units affect log1p. | Hodson (2022), transformed errors; package log1p convention |
| RMSLE (`rmsle`) | $\mathrm{RMSLE}=\sqrt{\frac{1}{n}\sum_{i=1}^{n}[\log(1+o_i)-\log(1+p_i)]^2}$ | Nonnegative; 0 ideal. Same domain as MSLE. | Hodson (2022); package log1p convention |
| RAE (`rae`) | $\mathrm{RAE}=\frac{\sum_{i=1}^{n}\lvert o_i-p_i\rvert}{\sum_{i=1}^{n}\lvert o_i-\bar o\rvert}$ | Nonnegative; 0 ideal, 1 observed-mean absolute-error benchmark; NA if A=0. | Hyndman & Koehler (2006); mean benchmark |
| RRMSE (`rrmse`) | $\mathrm{RRMSE}=100\frac{\mathrm{RMSE}}{\lvert\bar o\rvert}$ | Nonnegative percent; 0 ideal; NA at zero mean, unstable near zero. | Willmott et al. (1985), RMSE; package absolute-mean normalization |
| Willmott's $d$ (`willmott_d`) | $d=1-\frac{\sum_{i=1}^{n}(o_i-p_i)^2}{\sum_{i=1}^{n}(\lvert p_i-\bar o\rvert+\lvert o_i-\bar o\rvert)^2}$ | [0,1]; 1 ideal. NA if D=0 (identical constant vectors); sensitive to large errors. | Willmott et al. (1985) |
| KGE (`kge`) | $\mathrm{KGE}=1-\sqrt{(r-1)^2+\left(\frac{s_p}{s_o}-1\right)^2+\left(\frac{\bar p}{\bar o}-1\right)^2}$ | At most 1; 1 ideal. Zero is not NSE's benchmark. NA for zero observation mean or either constant vector. | Gupta et al. (2009) |

### Redundancy and comparison conventions

Each statistic now appears once: use `bias`, `mae`, `rmse`, `correlation`,
`R2`, and `ccc` instead of the former duplicate columns `ME`, `MAE`, `RMSE`,
`r`, `nse`/`NSE`/`MEC`, and `rhoC`. Standalone `nse()` and `mec()` remain
aliases for `R2()`. `Cb` is retained because it measures a separate component.
KGE is included in extended output. Quantile loss is requested separately
with `pinball_loss()`, since it requires a scientifically chosen quantile.

Related metrics are not independent evidence: MSE is RMSE squared, RPD is
the reciprocal of NRMSE when defined, and SEP equals cRMSE times
$\sqrt{n/(n-1)}$. These have different units or established interpretations
and remain available. With identical pairs,
$R^2=1-n\,\mathrm{NRMSE}^2/(n-1)$ and
$\mathrm{RMSE}^2=b^2+\mathrm{cRMSE}^2$.

Missing pairs are removed separately for each model by default; use a shared
complete validation set for fair comparisons. With `na.rm = FALSE`, missing
pairs produce missing metrics. Core correlation, efficiency, SD ratio, CCC,
and Cb require at least two pairs and varying observations. Percentage errors
need a meaningful zero and are unstable with small denominators. In particular,
the existing MAPE/sMAPE fraction scale is preserved, unlike MPE/RRMSE's percent
scale.

### References

* Legates, D. R. and McCabe, G. J. (1999). Evaluating the use of goodness-of-fit measures in hydrologic and hydroclimatic model validation. *Water Resources Research*, 35, 233-241. <https://doi.org/10.1029/1998WR900018>
* Willmott, C. J. and Matsuura, K. (2005). Advantages of the mean absolute error (MAE) over the root mean square error (RMSE) in assessing average model performance. *Climate Research*, 30, 79-82. <https://doi.org/10.3354/cr030079>
* Hodson, T. O. (2022). Root-mean-square error (RMSE) or mean absolute error (MAE): when to use them or not. *Geoscientific Model Development*, 15, 5481-5487. <https://doi.org/10.5194/gmd-15-5481-2022>
* Taylor, K. E. (2001). Summarizing multiple aspects of model performance in a single diagram. *Journal of Geophysical Research*, 106, 7183-7192. <https://doi.org/10.1029/2000JD900719>
* Nash, J. E. and Sutcliffe, J. V. (1970). River flow forecasting through conceptual models part I. *Journal of Hydrology*, 10, 282-290. <https://doi.org/10.1016/0022-1694(70)90255-6>
* Janssen, P. H. M. and Heuberger, P. S. C. (1995). Calibration of process-oriented models. *Ecological Modelling*, 83, 55-66. <https://doi.org/10.1016/0304-3800(95)00084-9>
* Lin, L. I.-K. (1989). A concordance correlation coefficient to evaluate reproducibility. *Biometrics*, 45, 255-268. <https://doi.org/10.2307/2532051>
* Bellon-Maurel, V., Fernandez-Ahumada, E., Palagos, B., Roger, J.-M., and McBratney, A. (2010). Critical review of chemometric indicators commonly used for assessing the quality of the prediction of soil attributes by NIR spectroscopy. *Trends in Analytical Chemistry*, 29, 1073-1081. <https://doi.org/10.1016/j.trac.2010.05.006>
* Hyndman, R. J. and Koehler, A. B. (2006). Another look at measures of forecast accuracy. *International Journal of Forecasting*, 22, 679-688. <https://doi.org/10.1016/j.ijforecast.2006.03.001>
* Willmott, C. J., Ackleson, S. G., Davis, R. E., Feddema, J. J., Klink, K. M., Legates, D. R., O'Donnell, J., and Rowe, C. M. (1985). Statistics for the evaluation and comparison of models. *Journal of Geophysical Research*, 90, 8995-9005. <https://doi.org/10.1029/JC090iC05p08995>
* Gupta, H. V., Kling, H., Yilmaz, K. K., and Martinez, G. F. (2009). Decomposition of the mean squared error and NSE performance criteria. *Journal of Hydrology*, 377, 80-91. <https://doi.org/10.1016/j.jhydrol.2009.08.003>

## Next steps

Use the [uncertainty-validation article](predictive-uncertainty.html) when
models provide intervals or predictive distributions, and the
[summary-diagrams article](summary-diagrams.html) for graphical comparison of
multiple models.
