---
title: "Point-prediction evaluation"
author: "Alexandre M.J.-C. Wadoux"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Point-prediction evaluation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
```

Continuous predictions can come from machine learning, statistical models, or
physical models. `modelskill` evaluates the paired predictions and observations,
not the algorithm that produced them. Use an independent validation set or an
appropriate resampling scheme.

```{r data}
library(modelskill)
set.seed(123)
obs <- seq(0, 10, length.out = 100) + rnorm(100, sd = 1)
models <- list(
  Good = obs + rnorm(100, sd = 0.5),
  Biased = obs + 1,
  Noisy = obs + rnorm(100, sd = 2)
)
```

## Start with observed versus predicted values

```{r scatter}
plot_data <- data.frame(
  observed = rep(obs, times = length(models)),
  predicted = unlist(models, use.names = FALSE),
  model = rep(names(models), each = length(obs))
)
ggplot2::ggplot(plot_data, ggplot2::aes(observed, predicted, colour = model)) +
  ggplot2::geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  ggplot2::geom_point(alpha = 0.7, size = 1.5) +
  ggplot2::coord_equal() + ggplot2::theme_classic() +
  ggplot2::labs(x = "Observed", y = "Predicted", colour = "Model")
```

The equality line represents perfect predictions. The `Good` model stays close
to it, `Biased` is shifted upward, and `Noisy` is more dispersed.

## Complementary metrics

```{r metrics}
model_metrics(models, obs)
```

No single metric is sufficient. Mean error (ME) describes signed systematic
error under the package convention `obs - pred`; negative ME indicates
overprediction. MAE and RMSE describe error magnitude in response units; RMSE
gives greater weight to large errors. Pearson correlation describes pattern
association, not agreement in level.

`nse()`, `mec()`, and uppercase `R2()` are aliases for the same
mean-benchmark efficiency coefficient. Lowercase `r2()` is squared Pearson
correlation and must not be interpreted as the efficiency coefficient.
Concordance (`ccc()`) combines correlation with level and spread agreement.

For specialised workflows, request robust, relative, percentage, and quantile
loss metrics explicitly:

```{r extended}
model_metrics(models, obs, extended = TRUE)
```

## Next steps

Use the [uncertainty-validation article](predictive-uncertainty.html) when
models provide intervals or predictive distributions, and the
[summary-diagrams article](summary-diagrams.html) for graphical comparison of
multiple models.
