---
title: "Predictive-uncertainty evaluation"
author: "Alexandre M.J.-C. Wadoux"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Predictive-uncertainty evaluation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
```

This article evaluates *quantified predictive uncertainty*. The same principles
apply to uncertainty estimates from machine learning (such as quantile
regression, ensembles, and conformal intervals), Bayesian models,
geostatistics, and process-based models.

Two properties should be assessed together:

* **Calibration** (or reliability): stated probabilities agree with empirical
  frequencies.
* **Sharpness**: intervals or distributions are sufficiently concentrated.
  Sharpness is valuable only when calibration is adequate.

```{r data}
library(modelskill)
set.seed(123)
obs <- seq(0, 10, length.out = 100) + rnorm(100, sd = 1)
pred <- obs + rnorm(100, sd = 0.5)
predictive_sd <- rep(0.8, length(obs))
lower95 <- pred - 1.96 * predictive_sd
upper95 <- pred + 1.96 * predictive_sd
```

## Prediction intervals

```{r intervals}
uncertainty_metrics(obs, lower = lower95, upper = upper95, level = .95)
```

PICP is between 0 and 1 and is ideal when it equals the nominal level. PICP
error is ideal at zero: negative values indicate under-coverage, while positive
values indicate over-coverage. PIW is average interval width in response units;
smaller means sharper intervals, conditional on good calibration. Interval
score combines width and penalties for misses; lower is better and scores are
comparable only for the same data, response scale, and interval level.

## Reliability across interval levels

```{r coverage}
gg_coverage(obs, pred = pred, predictive_sd = predictive_sd)
accuracy_plot_metrics(obs, pred = pred, predictive_sd = predictive_sd)
```

This PICP reliability plot is also called an *accuracy plot* in geostatistical
uncertainty validation. A curve near the 1:1 line is desirable. Its area-based
summaries have ideal value zero.

## Quantiles, PIT, and proper scores

```{r distribution}
qhat <- cbind(`0.1` = pred - 1.282 * predictive_sd,
              `0.5` = pred,
              `0.9` = pred + 1.282 * predictive_sd)
gg_qcp(obs, qhat, c(.1, .5, .9))
gg_pit(pit(stats::pnorm(obs, mean = pred, sd = predictive_sd)))
crps(obs, pred = pred, predictive_sd = predictive_sd)
```

At quantile probability \(p\), QCP is ideal when it equals \(p\). PIT values
should be uniformly distributed across validation cases. CRPS jointly rewards
calibration and sharpness; zero is ideal and lower values are better. For
predictive samples, `crps_decomposition()` reports the non-negative RELI
component, ideal at zero. `log_score()` is also lower-is-better but strongly
penalizes observations assigned very low density.

Schmidinger, J. and Heuvelink, G. B. M. (2023). Validation of uncertainty
predictions in digital soil mapping. *Geoderma*, 437, 116585.
<https://doi.org/10.1016/j.geoderma.2023.116585>
