---
title: "Predictive-uncertainty evaluation"
author: "Alexandre M.J.-C. Wadoux"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Predictive-uncertainty evaluation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)

```

## 1. Why validate predictive uncertainty?

Point-prediction metrics assess how close predictions are to observations, but
many models also quantify how uncertain each prediction is. This information may
be supplied as prediction intervals, predicted quantiles, a predictive mean and
standard deviation, or a complete predictive distribution.

These probabilistic predictions should be validated just like point predictions.
A poorly calibrated uncertainty estimate can give users unwarranted confidence
in a prediction even when the point prediction itself appears reasonable.

Two properties are particularly important:

- **Calibration**, also called **reliability**, describes whether stated
  probabilities agree with observed frequencies. For example, approximately
  90% of independent observations should fall inside intervals presented as
  90% prediction intervals.
- **Sharpness** describes how concentrated the predictive distributions are.
  Narrower prediction intervals correspond to sharper predictions.

Sharpness should always be interpreted conditional on adequate calibration. Very
narrow intervals are not useful when observations frequently fall outside them.
The general objective is therefore to obtain predictive distributions that are
as sharp as possible while remaining well calibrated
[@schmidinger2023; @gneitingraftery2007].

As with point-prediction validation, these diagnostics should be calculated
using independent validation data or an appropriate resampling strategy.

## 2. Which form of predictive uncertainty do you have?

Probabilistic predictions can be supplied in several forms, including prediction
intervals, predicted quantiles, predictive means and standard deviations, and
full predictive distributions. This tutorial develops two broad worked cases:

- **Case A — predictive mean and standard deviation**: the predictive
  distribution is summarized by its mean and standard deviation. Prediction
  intervals, quantiles, CDF values, and densities can be derived when an
  appropriate distributional assumption is available.
- **Case B — full predictive distribution**: the predictive distribution is
  represented directly by predictive samples or draws. This can preserve
  features such as skewness, heavy tails, and multimodality that may not be
  represented by a mean and standard deviation alone.

Prediction intervals and quantiles may also be supplied directly by a model.
The same validation principles apply regardless of how they were generated. A
model may also provide several representations of the same predictive
distribution.

## 3. Functions used in this tutorial

The main `modelskill` functions used below are:

| Function | Description | Reference |
| :-- | :-- | :-- |
| [`picp()`](../reference/picp.html)                                   | Calculates the prediction interval coverage probability, that is, the proportion of observations falling within a prediction interval. | [@goovaerts2001; @schmidinger2023]            |
| [`coverage_error()`](../reference/coverage_error.html)               | Calculates the difference between empirical and nominal prediction interval coverage.                                                  | [@schmidinger2023]                              |
| [`interval_width()`](../reference/interval_width.html) | Calculates the average width of prediction intervals and therefore summarizes their sharpness. | [@schmidinger2023] |
| [`interval_score()`](../reference/interval_score.html)               | Calculates a proper scoring rule that combines prediction interval width with penalties for observations outside the interval.         | [@gneitingraftery2007]                                   |
| [`uncertainty_metrics()`](../reference/uncertainty_metrics.html)     | Calculates PICP, coverage error, interval width, and interval score together.                                                          | [@gneitingraftery2007; @schmidinger2023] |
| [`gg_coverage()`](../reference/gg_coverage.html)                     | Produces a reliability plot comparing nominal prediction interval coverage with empirical coverage across several interval levels.     | [@goovaerts2001; @schmidinger2023]            |
| [`accuracy_plot_metrics()`](../reference/accuracy_plot_metrics.html) | Summarizes deviations between the empirical coverage curve and the ideal 1:1 reliability line.                                         | [@goovaerts2001; @schmidinger2023]            |
| [`qcp()`](../reference/qcp.html)                                     | Calculates quantile coverage probability, the proportion of observations below each predicted quantile.                                | [@schmidinger2023]                              |
| [`gg_qcp()`](../reference/gg_qcp.html)                               | Produces a quantile calibration plot comparing nominal quantile probabilities with empirical quantile coverage.                        | [@schmidinger2023]                              |
| [`pit()`](../reference/pit.html) | Validates PIT values supplied as predictive CDF values evaluated at the observations. | [@gneiting2007; @schmidinger2023] |
| [`gg_pit()`](../reference/gg_pit.html)                               | Produces a PIT histogram for assessing the calibration of complete predictive distributions.                                           | [@gneitingraftery2007; @schmidinger2023] |
| [`crps()`](../reference/crps.html)                                   | Calculates the continuous ranked probability score, a proper score that evaluates the full predictive distribution.                    | [@hersbach2000; @gneitingraftery2007]                  |
| [`median_crps()`](../reference/median_crps.html)                     | Calculates the median case-wise CRPS, providing a robust summary of typical probabilistic prediction performance.                      | [@gneitingraftery2007]                                   |
| [`crps_decomposition()`](../reference/crps_decomposition.html)       | Decomposes CRPS into reliability and potential CRPS components when predictive samples are available.                                  | [@hersbach2000; @schmidinger2023]             |
| [`log_score()`](../reference/log_score.html)                         | Calculates the logarithmic score from predictive density values evaluated at the observations.                                         | [@gneitingraftery2007]                                   |

The appropriate functions depend on the available model output. For example,
`picp()`, `coverage_error()`, `interval_width()`, and `interval_score()` require
prediction-interval bounds. `crps()` can use either a normal predictive
distribution represented by a predictive mean and standard deviation or a
matrix of predictive samples. `crps_decomposition()` is intended for equally
weighted predictive samples, while `pit()` and `log_score()` require predictive
CDF or density values evaluated at the observations.

## 4. Example data

We first create a simple situation in which the predictive distributions are
correctly calibrated by construction.

```{r}
library(modelskill)

set.seed(123)

n <- 500

# Predictive means
pred <- seq(0, 10, length.out = n)

# True predictive standard deviation
predictive_sd <- rep(1, n)

# Independent observations generated from the predictive distributions
obs <- stats::rnorm(
  n,
  mean = pred,
  sd = predictive_sd
)

```

Here, observation \(i\) is generated from

\[
Y_i \sim N(\mathrm{pred}_i, \sigma_i^2),
\]

with \(\sigma_i = 1\). Because the predictive model and data-generating process
are the same, the uncertainty estimates should be approximately calibrated.
Departures from perfect calibration still occur because validation is based on
a finite sample.

## 5. Case A: predictive mean and standard deviation

A common output from geostatistical models, Bayesian approximations, Gaussian
processes, and other statistical models is a predictive mean together with a
predictive standard deviation.

If a normal predictive distribution is appropriate, prediction intervals,
quantiles, CDF values, and predictive densities can all be obtained from these
two quantities.

### 5.1 Prediction intervals

For a central 95% prediction interval:

```{r}
lower95 <- pred + stats::qnorm(0.025) * predictive_sd
upper95 <- pred + stats::qnorm(0.975) * predictive_sd

```

A 95% prediction interval means that, under calibration, approximately 95% of
independent observations should fall between the corresponding lower and upper
bounds.

#### 5.1.1 Prediction interval coverage probability

The **prediction interval coverage probability (PICP)** is the empirical
proportion of observations contained in the interval:

\[
\mathrm{PICP}(\tau) =
\frac{1}{n}
\sum_{i=1}^{n}
I(l_i \leq y_i \leq u_i).
\]

where \(\tau\) is the nominal interval level.

```{r}
picp(obs, lower95, upper95)

```

For a 95% prediction interval, a well-calibrated model should return a value
close to 0.95.

`modelskill` reports PICP on the probability scale from 0 to 1 rather than as a
percentage.

#### 5.1.2 Coverage error

[`coverage_error()`](../reference/coverage_error.html) directly compares
empirical and nominal coverage:

\[
\mathrm{coverage\ error} = \mathrm{PICP}(\tau) - \tau.
\]

```{r}
coverage_error(
  obs,
  lower95,
  upper95,
  level = 0.95
)

```

Interpretation is straightforward:

- **0**: empirical and nominal coverage agree;
- **negative**: under-coverage, meaning that observations fall outside the
  intervals too frequently;
- **positive**: over-coverage, meaning that observations fall inside the
  intervals more frequently than required.

Under-coverage often indicates that predictive uncertainty has been
underestimated, whereas substantial over-coverage can indicate unnecessarily
wide uncertainty estimates.

#### 5.1.3 Prediction interval width

Calibration alone is not sufficient. A model could achieve high coverage simply
by producing extremely wide intervals.

[`interval_width()`](../reference/interval_width.html) measures average
prediction interval width:

\[
\mathrm{PIW}(\tau) =
\frac{1}{n}
\sum_{i=1}^{n}
(u_i-l_i).
\]

```{r}
interval_width(
  obs,
  lower95,
  upper95
)

```

PIW has the same units as the response variable. Smaller values mean sharper
prediction intervals.

Importantly, PIW does not measure calibration and does not depend on where the
observations fall. It should therefore always be interpreted together with a
coverage diagnostic such as PICP.

#### 5.1.4 Interval score

The [`interval_score()`](../reference/interval_score.html) combines sharpness
and calibration in a single proper score.

```{r}
interval_score(
  obs,
  lower95,
  upper95,
  level = 0.95
)

```

The score contains the interval width plus penalties when an observation falls
below or above the interval. The farther an observation lies outside the
interval, the larger the penalty.

Lower interval scores are better. Scores should only be compared for the same
validation data, response scale, and interval level.

#### 5.1.5 Calculate all interval diagnostics together

[`uncertainty_metrics()`](../reference/uncertainty_metrics.html) is a
convenience wrapper that returns the principal interval diagnostics together:

```{r}
uncertainty_metrics(
  obs,
  lower = lower95,
  upper = upper95,
  level = 0.95
)

```

It returns:

- `picp`: empirical prediction interval coverage;
- `picp_error`: empirical minus nominal coverage;
- `interval_width`: average interval width;
- `interval_score`: proper interval score.

The individual functions remain useful when only one diagnostic is required.

### 5.2 Why calibration and sharpness must be considered together

Consider a second uncertainty estimate that uses the same predictive means but
standard deviations that are only half as large:

```{r}
sd_too_small <- rep(0.5, n)

lower95_narrow <- pred + stats::qnorm(0.025) * sd_too_small
upper95_narrow <- pred + stats::qnorm(0.975) * sd_too_small

uncertainty_metrics(
  obs,
  lower = lower95_narrow,
  upper = upper95_narrow,
  level = 0.95
)

```

These intervals are much narrower and therefore appear sharper. However, their
coverage should be substantially below 0.95 because the predictive uncertainty
has been underestimated.

Compare this with the calibrated intervals:

```{r}
uncertainty_metrics(
  obs,
  lower = lower95,
  upper = upper95,
  level = 0.95
)

```

This illustrates why a smaller interval width is desirable only when
calibration remains adequate.

### 5.3 Reliability across multiple interval levels

A single PICP assesses one interval level only. It is often more informative to
evaluate calibration across many central prediction intervals.

[`gg_coverage()`](../reference/gg_coverage.html) constructs this reliability
curve directly from predictive means and standard deviations:

```{r}
gg_coverage(
  obs,
  pred = pred,
  predictive_sd = predictive_sd
)

```

Nominal interval coverage is shown on the horizontal axis and empirical PICP on
the vertical axis.

A well-calibrated model should follow the 1:1 line.

- Points **below** the line indicate under-coverage and generally
  underestimated uncertainty.
- Points **above** the line indicate over-coverage and generally overly
  conservative uncertainty.

The same plot for the deliberately underestimated predictive standard
deviations shows the difference:

```{r}
gg_coverage(
  obs,
  pred = pred,
  predictive_sd = sd_too_small
)

```

In geostatistical uncertainty validation this display has traditionally also
been called an **accuracy plot** [@goovaerts2001], although **reliability**
**plot** is the more general terminology used for probabilistic predictions.

#### 5.3.1 Numerical summaries of the reliability plot

The reliability or accuracy plot can be interpreted visually, but departures
from the ideal 1:1 line can also be summarized numerically. Accuracy plots were
introduced for the assessment of local uncertainty by @deutsch1997 and were
subsequently used for geostatistical uncertainty evaluation in soil science by
@goovaerts2001. The same framework, including numerical summaries of
departure from the 1:1 line, was used by @wadoux2018.

For a nominal central prediction-interval level \(p\), let
\(\mathrm{PICP}(p)\) denote the empirical proportion of validation observations
contained in the corresponding prediction intervals. If predictive uncertainty
is well calibrated,

\[
\mathrm{PICP}(p) \approx p
\]

over the range of evaluated probabilities. The empirical coverage curve should
therefore follow the 1:1 line.

[`accuracy_plot_metrics()`](../reference/accuracy_plot_metrics.html) summarizes
the magnitude and direction of departures from this line:

```{r}
accuracy_plot_metrics(
  obs,
  pred = pred,
  predictive_sd = predictive_sd
)
```

For the deliberately underestimated predictive uncertainty:

```{r}
accuracy_plot_metrics(
  obs,
  pred = pred,
  predictive_sd = sd_too_small
)
```

The principal summary is `absolute_deviation`, which corresponds to the total
area between the empirical coverage curve and the ideal 1:1 line:

\[
A =
\int_0^1
\left|
\mathrm{PICP}(p)-p
\right|
\,dp.
\]

A value of \(A=0\) indicates perfect calibration. Increasing values indicate
greater overall disagreement between nominal and empirical coverage across the
evaluated interval levels. This measure summarizes the magnitude of
miscalibration but does not, by itself, indicate whether predictive uncertainty
is predominantly overestimated or underestimated.

The total deviation can therefore be separated into the area above the 1:1
line,

\[
A_{\mathrm{over}} =
\int_0^1
\max\left\{
\mathrm{PICP}(p)-p,0
\right\}
\,dp,
\]

and the area below the line,

\[
A_{\mathrm{under}} =
\int_0^1
\max\left\{
p-\mathrm{PICP}(p),0
\right\}
\,dp.
\]

These quantities are returned as:

- `absolute_deviation`: the total area \(A\) between the empirical coverage
  curve and the 1:1 line. Smaller values indicate better overall calibration,
  with zero being ideal;
- `over_uncertainty`: the area \(A_{\mathrm{over}}\) above the 1:1 line. In
  this region empirical coverage is greater than nominal coverage, indicating
  over-coverage and, in general, predictive uncertainty that is too large;
- `under_uncertainty`: the area \(A_{\mathrm{under}}\) below the 1:1 line. In
  this region empirical coverage is smaller than nominal coverage, indicating
  under-coverage and, in general, predictive uncertainty that is too small;
- `over_percent`: the percentage of the total absolute deviation occurring
  above the 1:1 line;
- `under_percent`: the percentage of the total absolute deviation occurring
  below the 1:1 line.

The last two quantities correspond to the relative contributions of
overestimation and underestimation to the total calibration error:

\[
P_{\mathrm{over}} =
\frac{A_{\mathrm{over}}}{A},
\qquad
P_{\mathrm{under}} =
\frac{A_{\mathrm{under}}}{A}.
\]

`accuracy_plot_metrics()` reports these quantities as percentages. When
`absolute_deviation` is greater than zero,

\[
\mathrm{over\_percent} + \mathrm{under\_percent} = 100.
\]

These percentages should not be interpreted as percentages of observations.
They describe how the total *area of miscalibration* is partitioned above and
below the 1:1 line.

For example, a large `under_percent` indicates that most of the departure from
ideal calibration results from empirical coverage being lower than nominal
coverage. The prediction intervals are therefore generally too narrow, and
predictive uncertainty is underestimated. Conversely, a large `over_percent`
indicates that most of the departure results from empirical coverage exceeding
nominal coverage, which is generally associated with prediction intervals that
are too wide and predictive uncertainty that is overestimated.

These summaries quantify **calibration**, not **sharpness**. A predictive model
may achieve good coverage while producing unnecessarily wide prediction
intervals. Calibration summaries should therefore be interpreted together with
measures of interval width or proper scoring rules when comparing probabilistic
predictions.

The area summaries are most informative when the reliability curve is evaluated
over a sufficiently dense range of nominal interval levels. With only one or a
few interval levels, a single coverage error is usually easier to interpret than
an integrated area.

The interpretation above follows the accuracy-plot framework of [@deutsch1997;
@goovaerts2001], with the absolute-deviation and directional summaries
described and applied in @wadoux2018.

### 5.4 Why PICP alone is not enough

PICP measures whether observations fall inside an interval, but it does not
describe **where** the observations that fall outside the interval occur.

This means that PICP can hide a **one-sided bias**.

For example, for a central 90% prediction interval we expect approximately 5%
of observations below the lower boundary and 5% above the upper boundary.
A model could instead have approximately 10% below the lower boundary and
almost none above the upper boundary while still obtaining approximately 90%
total coverage.

@schmidinger2023 showed that this limitation can occur in
practice and recommended complementing PICP with quantile-based calibration or
PIT diagnostics.

We can reproduce the idea with a deliberately shifted predictive distribution.

The predictive mean is shifted upward by 0.5 response units. Its predictive
standard deviation is also increased so that central 90% coverage can remain
close to its nominal value.

```{r}
biased_pred <- pred + 0.5
biased_sd <- rep(1.118, n)

lower90_biased <-
  biased_pred + stats::qnorm(0.05) * biased_sd

upper90_biased <-
  biased_pred + stats::qnorm(0.95) * biased_sd

uncertainty_metrics(
  obs,
  lower = lower90_biased,
  upper = upper90_biased,
  level = 0.90
)

```

The total PICP may look acceptable even though the predictive distribution is
systematically displaced. Individual quantiles reveal this problem more
clearly.

### 5.5 Quantile coverage probability

Prediction interval coverage probability evaluates whether observations fall
inside prediction intervals, but it does not distinguish between the lower and
upper tails. Quantile coverage probability (QCP) provides a complementary
calibration diagnostic by evaluating individual predictive quantiles.

For a correctly calibrated predictive quantile at probability \(p\),
approximately a fraction \(p\) of observations should fall below that predicted
quantile. The **quantile coverage probability (QCP)** is

\[
\mathrm{QCP}(p) =
\frac{1}{n}
\sum_{i=1}^{n}
I(y_i \leq q_{i,p}),
\]

where \(q_{i,p}\) is the predicted \(p\)-quantile for observation \(i\).

For a calibrated predictive distribution,

- a predicted 0.05 quantile should have QCP close to 0.05;
- a predicted 0.50 quantile should have QCP close to 0.50;
- a predicted 0.95 quantile should have QCP close to 0.95.

It is generally more informative to evaluate QCP across several quantile levels
rather than at only a few selected probabilities. We therefore evaluate
quantiles from 0.05 to 0.95 at increments of 0.05.

```{r}
q_levels <- seq(0.05, 0.95, by = 0.05)

qhat <- vapply(
  q_levels,
  function(p) {
    pred + stats::qnorm(p) * predictive_sd
  },
  numeric(n)
)

qcp(
  obs,
  quantiles = qhat,
  levels = q_levels
)
```

For a well-calibrated predictive distribution, the empirical QCP values should
be close to their corresponding nominal quantile probabilities.

[`gg_qcp()`](../reference/gg_qcp.html) displays this relationship graphically:

```{r}
gg_qcp(
  obs,
  quantiles = qhat,
  levels = q_levels
)
```

The horizontal axis gives the nominal predictive quantile probability and the
vertical axis gives its empirical QCP. Points close to the 1:1 line indicate
well-calibrated predictive quantiles.

When the predictive distribution is represented by a predictive mean and
standard deviation and a normal distribution is appropriate, `gg_qcp()` can
also generate these quantiles directly:

```{r}
gg_qcp(
  obs,
  pred = pred,
  predictive_sd = predictive_sd
)
```

This differs from [`gg_coverage()`](../reference/gg_coverage.html). The
reliability or accuracy plot produced by `gg_coverage()` evaluates the
**joint coverage of central prediction intervals**, whereas `gg_qcp()` evaluates
**individual predictive quantiles**. For example, a central 90% prediction
interval jointly evaluates its 0.05 and 0.95 quantile bounds. QCP instead
evaluates those two quantiles separately.

This distinction is important because satisfactory central interval coverage
does not necessarily imply that both tails are calibrated. A prediction
interval may contain approximately the expected proportion of observations even
when too many observations fall below its lower bound and too few fall above its
upper bound.

Now examine the deliberately biased predictive distribution:

```{r}
qhat_biased <- vapply(
  q_levels,
  function(p) {
    biased_pred + stats::qnorm(p) * biased_sd
  },
  numeric(n)
)

qcp(
  obs,
  quantiles = qhat_biased,
  levels = q_levels
)

gg_qcp(
  obs,
  quantiles = qhat_biased,
  levels = q_levels
)
```

The same plot can be generated directly from its predictive mean and standard
deviation:

```{r}
gg_qcp(
  obs,
  pred = biased_pred,
  predictive_sd = biased_sd
)
```

Departures from the 1:1 line indicate quantile miscalibration. QCP above its
nominal probability means that observations fall below the predicted quantile
more frequently than expected, so the predicted quantile tends to be too high.
QCP below its nominal probability means that observations fall below the
predicted quantile less frequently than expected, so the predicted quantile
tends to be too low.

QCP is therefore particularly useful for identifying asymmetric or one-sided
miscalibration that may remain hidden when central prediction-interval coverage
is considered alone.

### 5.6 Probability integral transform

When a complete predictive cumulative distribution function \(F_i\) is
available, calibration can also be assessed through the **probability integral
transform (PIT)**:

\[
u_i = F_i(y_i).
\]

For a calibrated continuous predictive distribution, PIT values should be
uniformly distributed between 0 and 1.

For the normal predictive distributions in this example, evaluate each CDF at
its corresponding observation:

```{r}
pit_normal <- pit(
  stats::pnorm(
    obs,
    mean = pred,
    sd = predictive_sd
  )
)

gg_pit(pit_normal)

```

There is no ideal PIT value for an individual observation. Calibration is
evaluated from the distribution of PIT values across the validation set.

A roughly flat histogram is desirable.

Common deviations are informative:

- excess observations near **0 and 1** generally indicate predictive
  distributions that are too narrow;
- excess observations near the **centre** generally indicate distributions that
  are too wide;
- systematic asymmetry or a slope can indicate directional bias.

For example, PIT immediately reveals the uncertainty estimates that were made
too narrow:

```{r}
pit_narrow <- pit(
  stats::pnorm(
    obs,
    mean = pred,
    sd = sd_too_small
  )
)

gg_pit(pit_narrow)

```

The biased distribution can also be inspected:

```{r}
pit_biased <- pit(
  stats::pnorm(
    obs,
    mean = biased_pred,
    sd = biased_sd
  )
)

gg_pit(pit_biased)

```

PIT and QCP convey closely related calibration information but emphasize
different aspects visually. @schmidinger2023 recommend them as
useful complements to PICP, particularly because they can expose calibration
problems that central interval coverage alone may hide.

### 5.7 Proper scoring rules

Calibration diagnostics such as PICP, QCP, and PIT ask whether a predictive
distribution is statistically reliable.

When several probabilistic models must be compared, it is also useful to have a
single numerical criterion that rewards both calibration and sharpness.
**Proper scoring rules** are designed for this purpose.

A proper scoring rule does not reward a model for deliberately reporting a
predictive distribution different from the distribution it actually represents
[@gneitingraftery2007].

#### 5.7.1 Continuous ranked probability score

The **continuous ranked probability score (CRPS)** compares the complete
predictive CDF with the realised observation.

[`crps()`](../reference/crps.html) supports two important input formats in
`modelskill`.

For a normal predictive distribution:

```{r}
crps(
  obs,
  pred = pred,
  predictive_sd = predictive_sd
)

```

CRPS has the same units as the response variable. Zero is ideal and lower
values are better.

Compare the calibrated and underestimated uncertainty models:

```{r}
crps(
  obs,
  pred = pred,
  predictive_sd = predictive_sd
)

crps(
  obs,
  pred = pred,
  predictive_sd = sd_too_small
)

```

CRPS simultaneously reflects the location of the predictive distribution and
its spread.

#### 5.7.2 Median CRPS

Mean CRPS can be influenced by a small number of very poor predictive
distributions.

[`median_crps()`](../reference/median_crps.html) summarizes the typical
case more robustly:

```{r}
median_crps(
  obs,
  pred = pred,
  predictive_sd = predictive_sd
)

```

Mean and median CRPS answer slightly different questions. Mean CRPS is usually
preferred for overall probabilistic model comparison, whereas median CRPS can
help determine whether the mean score is dominated by a few extreme cases.

#### 5.7.3 Logarithmic score

The [`log_score()`](../reference/log_score.html) is another proper score. It
uses the predictive density assigned to the observed value:

\[
\mathrm{Log\ score} =
-\frac{1}{n}
\sum_{i=1}^{n}
\log f_i(y_i).
\]

For the normal predictive distributions:

```{r}
density_at_obs <- stats::dnorm(
  obs,
  mean = pred,
  sd = predictive_sd
)

log_score(
  obs,
  density_at_obs
)

```

Lower values are better.

The logarithmic score strongly penalizes observations that were assigned very
low predictive density. It can therefore be particularly sensitive to failures
in the tails of the predictive distribution.

## 6. Case B: full predictive distributions supplied as samples

Many probabilistic models do not return only a predictive mean and standard
deviation. Instead, they provide a set of possible values from the predictive
distribution for every validation case.

Typical examples include:

- posterior predictive draws from Bayesian models;
- simulation-based models;
- ensemble predictions;
- bootstrap predictive distributions;
- some machine-learning methods;
- Monte Carlo uncertainty propagation.

`modelskill` represents this type of prediction as a matrix with:

- one **row** per validation observation;
- one **column** per predictive draw.

We generate 200 predictive draws for each observation:

```{r}
set.seed(456)

n_draws <- 200

predictive_samples <- sapply(
  seq_len(n_draws),
  function(j) {
    stats::rnorm(
      n,
      mean = pred,
      sd = predictive_sd
    )
  }
)

dim(predictive_samples)

```

Each row of `predictive_samples` now represents an empirical approximation to
the predictive distribution for one observation.

#### 6.1 Prediction intervals from predictive samples

Prediction intervals can be obtained from the appropriate quantiles of the
predictive samples.

For a 95% central interval:

```{r}
lower95_samples <- apply(
  predictive_samples,
  1,
  stats::quantile,
  probs = 0.025,
  names = FALSE
)

upper95_samples <- apply(
  predictive_samples,
  1,
  stats::quantile,
  probs = 0.975,
  names = FALSE
)

uncertainty_metrics(
  obs,
  lower = lower95_samples,
  upper = upper95_samples,
  level = 0.95
)

```

The interpretation of PICP, interval width, coverage error, and interval score
is exactly the same as when the intervals were derived from a predictive mean
and standard deviation.

The validation tools depend on the resulting predictive uncertainty, not on the
algorithm used to generate it.

### 6.2 Reliability across interval levels from predictive samples

When only predictive samples are available, central prediction intervals can be
constructed for several nominal levels.

```{r}
interval_levels <- seq(0.10, 0.90, by = 0.10)

lower_list <- lapply(
  interval_levels,
  function(level) {
    alpha <- (1 - level) / 2

    apply(
      predictive_samples,
      1,
      stats::quantile,
      probs = alpha,
      names = FALSE
    )
  }
)

upper_list <- lapply(
  interval_levels,
  function(level) {
    alpha <- (1 - level) / 2

    apply(
      predictive_samples,
      1,
      stats::quantile,
      probs = 1 - alpha,
      names = FALSE
    )
  }
)

names(lower_list) <- sprintf("%.2f", interval_levels)
names(upper_list) <- sprintf("%.2f", interval_levels)

```

These lists can be supplied directly to `gg_coverage()`:

```{r}
gg_coverage(
  obs,
  lower = lower_list,
  upper = upper_list
)

```

and to `accuracy_plot_metrics()`:

```{r}
accuracy_plot_metrics(
  obs,
  lower = lower_list,
  upper = upper_list
)

```

Thus, `gg_coverage()` can be used both when uncertainty is represented by
normal predictive means and standard deviations and when intervals have been
derived from arbitrary predictive distributions.

### 6.3 Quantile calibration from predictive samples

Predicted quantiles can also be extracted from each predictive distribution.

```{r}
qhat_samples <- t(
  apply(
    predictive_samples,
    1,
    stats::quantile,
    probs = q_levels,
    names = FALSE
  )
)

qcp(
  obs,
  quantiles = qhat_samples,
  levels = q_levels
)

```

Plot the calibration:

```{r}
gg_qcp(
  obs,
  quantiles = qhat_samples,
  levels = q_levels
)

```

This approach does not require any parametric assumption about the shape of the
predictive distribution.

### 6.4 PIT from predictive samples

For an analytic distribution, PIT is calculated by evaluating the predictive
CDF at the observation.

With predictive samples, the predictive CDF can instead be approximated by the
fraction of predictive draws less than or equal to the observation:

```{r}
cdf_at_obs_samples <- rowMeans(
  sweep(
    predictive_samples,
    1,
    obs,
    FUN = "<="
  )
)

pit_samples <- pit(cdf_at_obs_samples)

gg_pit(pit_samples)

```

For calibrated predictive distributions, the resulting PIT histogram should
again be approximately uniform.

The difference is only how \(F_i(y_i)\) was obtained:

- with an analytic distribution, evaluate its CDF directly;
- with predictive samples, approximate the CDF empirically.

### 6.5 CRPS from predictive samples

`crps()` accepts the predictive sample matrix directly:

```{r}
crps(
  obs,
  distribution = predictive_samples
)

```

No normality assumption is required.

This is useful when predictive distributions are asymmetric, multimodal, or
otherwise poorly summarized by a mean and standard deviation.

The median case-wise CRPS is similarly available:

```{r}
median_crps(
  obs,
  distribution = predictive_samples
)

```

### 6.6 CRPS reliability decomposition

When equally weighted predictive samples are available,
[`crps_decomposition()`](../reference/crps_decomposition.html) decomposes CRPS
into a reliability component and potential CRPS:

```{r}
crps_decomposition(
  obs,
  distribution = predictive_samples
)

```

The output contains:

- `crps`: the total CRPS;
- `reliability`: the RELI component;
- `potential_crps`: the remaining component after removing reliability error.

The reliability component is non-negative and has an ideal value of zero.
It provides a numerical summary of distributional calibration.

@schmidinger2023 showed that RELI summarizes patterns also
visible in PICP, QCP, and PIT diagnostics, whereas the total CRPS additionally
reflects sharpness and predictive performance.

`crps_decomposition()` in `modelskill` is intended for equally weighted
predictive samples. It is therefore not called directly with a predictive mean
and standard deviation.

### 6.7 What about the logarithmic score with predictive samples?

`log_score()` requires the predictive density evaluated at each observation.

If a model provides an analytic density, calculate these density values and
supply them directly.

For example, in the normal case:

```{r}
density_at_obs <- stats::dnorm(
  obs,
  mean = pred,
  sd = predictive_sd
)

log_score(obs, density_at_obs)

```

Predictive samples alone do not uniquely define a smooth probability density.
A density-estimation method would first be required. `modelskill` therefore
does not automatically estimate a density from predictive samples before
calculating the logarithmic score.

## 7. Comparing the two input formats

The two examples above represent the same conceptual predictive uncertainty in
different ways.

| TaskMean + SDPredictive samples |                                |                                   |
| ------------------------------- | ------------------------------ | --------------------------------- |
| Construct intervals             | Normal quantiles               | Empirical quantiles               |
| `uncertainty_metrics()`         | Yes, after constructing bounds | Yes, after constructing bounds    |
| `gg_coverage()`                 | Directly                       | Via interval lists                |
| `qcp()` / `gg_qcp()`            | From normal quantiles          | From empirical quantiles          |
| `pit()` / `gg_pit()`            | From analytic CDF              | From empirical CDF                |
| `crps()`                        | Directly                       | Directly                          |
| `median_crps()`                 | Directly                       | Directly                          |
| `crps_decomposition()`          | No                             | Yes                               |
| `log_score()`                   | From analytic density          | Requires density estimation first |

The appropriate representation depends on what the prediction model provides.
There is generally no reason to reduce a rich predictive distribution to a mean
and standard deviation solely for validation if predictive samples or quantiles
are already available.

## 8. A practical workflow

There is no single statistic that completely evaluates predictive uncertainty.
Different diagnostics answer different questions.

### 8.1 If you only have prediction intervals

Start with:

```r
uncertainty_metrics(...)

```

and interpret PICP and interval width together. If intervals are available at
several nominal levels, also use:

```r
gg_coverage(...)
accuracy_plot_metrics(...)

```

### 8.2 If you have predicted quantiles

Use:

```r
qcp(...)
gg_qcp(...)

```

QCP is particularly useful because it can expose directional quantile bias that
may be hidden by central interval coverage.

### 8.3 If you have a predictive mean and standard deviation

When a normal predictive distribution is scientifically appropriate, you can
evaluate nearly the complete distribution using:

```r
gg_coverage(...)
pit(...)
gg_pit(...)
crps(...)
median_crps(...)
log_score(...)

```

Prediction intervals and quantiles can also be generated from the normal
distribution as shown above.

### 8.4 If you have full predictive samples

Retain them rather than reducing them unnecessarily to a mean and standard
deviation.

Use them to derive:

```r
# Intervals
uncertainty_metrics(...)

# Quantiles
qcp(...)
gg_qcp(...)

# Empirical CDF values
pit(...)
gg_pit(...)

# Proper scores
crps(...)
median_crps(...)
crps_decomposition(...)

```

This preserves asymmetry and other features of the predictive distribution.

## 9. Recommended interpretation

A practical uncertainty-validation analysis should generally combine several
complementary diagnostics.

1. **Check calibration.**
   Use PICP, QCP, PIT, or preferably more than one of these diagnostics.
2. **Check sharpness.**
   Examine prediction interval width, but only after verifying calibration.
3. **Inspect the whole distribution when possible.**
   QCP and PIT can reveal problems that a single central PICP cannot.
4. **Use proper scores when comparing models.**
   CRPS and interval score combine calibration and sharpness into numerical
   criteria suitable for relative model comparison.
5. **Do not select a model from one diagnostic alone.**
   Different metrics emphasize different aspects of probabilistic prediction
   quality and can therefore lead to different model rankings.

@schmidinger2023 demonstrated this explicitly for probabilistic
digital soil mapping: PICP could hide one-sided bias, whereas QCP and PIT
revealed it, and scoring rules provided additional information when comparing
competing probabilistic models.

