---
title: "Summary diagrams and diagnostic plots"
author: "Alexandre M.J.-C. Wadoux"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Summary diagrams and diagnostic plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 6)
```

Taylor, solar, and target diagrams provide complementary graphical summaries
for comparing continuous prediction models. They return ordinary `ggplot2`
objects and are most informative when interpreted alongside numerical metrics.
They deliberately show different aspects of predictive performance, so one
diagram should not be used as a substitute for the others.

```{r data}
library(modelskill)
set.seed(123)
obs <- seq(0, 10, length.out = 100) + rnorm(100, sd = 1)
models <- list(Good = obs + rnorm(100, sd = 0.5),
               Biased = obs + 1,
               Noisy = obs + rnorm(100, sd = 2))
```

## Taylor diagram

The Taylor diagram combines correlation, relative spread, and centred error.
Its radial coordinate is the standard-deviation ratio, its polar angle is the
inverse cosine of Pearson correlation, and its reference point represents a
perfectly matching prediction:

\[
\sigma^* = \frac{\mathrm{sd}(pred)}{\mathrm{sd}(obs)},\qquad
\theta = \arccos(r),\qquad (\sigma^*, r) = (1, 1).
\]

The distance from a model point to this reference is the standardized centred
root mean square difference (SDE\*):

\[
\mathrm{SDE}^* = \sqrt{1 + \sigma^{*2} - 2\sigma^*r}.
\]

`Good` should be near the reference point. `Noisy` should be farther away and
at a larger polar angle because it has weaker pattern agreement. A radius below
one indicates smoothing (too little variation); a radius above one indicates
too much variation. A Taylor diagram does **not** show mean error: `Biased`
can be near the reference point because its constant offset leaves correlation
and spread unchanged.

```{r taylor}
gg_taylor(models, obs, label = TRUE)
```

## Solar diagram

The solar diagram makes explicit the mean error omitted by a Taylor diagram.
It uses the standardized decomposition:

\[
\mathrm{RMSE}^{*2} = \mathrm{ME}^{*2} + \mathrm{SDE}^{*2}.
\]

Horizontal position is normalized mean error (ME\*) and vertical position is
SDE\*. Distance from the origin is RMSE\*. Under the package convention
`obs - pred`, points to the left overpredict and points to the right
underpredict. Thus `Biased` falls to the left, `Good` should be near the
origin, and `Noisy` should lie higher on the SDE\* axis.

The outer circle at \(\mathrm{RMSE}^*=1\) marks the mean-prediction benchmark:
points inside it have positive model efficiency (MEC/NSE/uppercase R2). The
pale regions are lower bounds on correlation, not exact correlation values.

```{r solar}
gg_solar(models, obs, label = TRUE, x.axis_begin = -1.5,
         x.axis_end = 1.5, y.axis_end = 2, by = 0.25)
```

## Target diagram

The target diagram uses the same RMSE* decomposition:

\[
\mathrm{RMSE}^{*2} = \mathrm{ME}^{*2} + \mathrm{SDE}^{*2}.
\]

It additionally gives SDE\* a sign according to whether prediction variability
is lower or higher than observed variability. The signed SDE coordinate thus
distinguishes smoother predictions from predictions that are too variable,
while ME* distinguishes overprediction from underprediction. Distance from the
origin remains RMSE\*, so `Good` should be nearest the origin. `Biased` is
chiefly displaced by mean error, whereas a model displaced along signed SDE\*
chiefly differs in variation or pattern.

The displayed axis titles preserve the original diagram orientation. Consult
the function documentation for the exact coordinate definitions when
interpreting horizontal and vertical position.

```{r target}
gg_target(models, obs, label = TRUE, axis_begin = -2, axis_end = 2, by = 0.5)
```

## Customisation

All three functions return normal ggplot objects. Their defaults retain the
scientific reference geometry and visual style, but standard ggplot additions
remain available:

```{r customise}
gg_taylor(models, obs) +
  ggplot2::labs(title = "Model comparison") +
  ggplot2::theme_minimal()
```

Wadoux, A. M. J.-C., Walvoort, D. J. J., and Brus, D. J. (2022). An
integrated approach for the evaluation of quantitative soil maps through
Taylor and solar diagrams. *Geoderma*, 405, 115332.
<https://doi.org/10.1016/j.geoderma.2021.115332>
