---
title: "Summary diagrams and diagnostic plots"
author: "Alexandre M.J.-C. Wadoux"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Summary diagrams and diagnostic plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 6)
```

Taylor, solar, and target diagrams provide complementary graphical summaries
for comparing continuous prediction models. They return ordinary `ggplot2`
objects and are most informative when interpreted alongside numerical metrics.

```{r data}
library(modelskill)
set.seed(123)
obs <- seq(0, 10, length.out = 100) + rnorm(100, sd = 1)
models <- list(Good = obs + rnorm(100, sd = 0.5),
               Biased = obs + 1,
               Noisy = obs + rnorm(100, sd = 2))
```

## Taylor diagram

The radial coordinate is the standard-deviation ratio
\(\sigma^*=\mathrm{sd}(pred)/\mathrm{sd}(obs)\), the polar angle is
\(\arccos(r)\), and the reference point is \((\sigma^*=1,r=1)\).
`Good` should lie closest to that reference. A constant offset does not affect
correlation or standard deviation, so `Biased` can still be near it.

```{r taylor}
gg_taylor(models, obs, label = TRUE)
```

## Solar diagram

The solar diagram displays standardized mean error (ME*) horizontally and
standardized centred error (SDE*) vertically. Under the `obs - pred`
convention, points to the left overpredict and points to the right underpredict.
Distance from the origin is standardized RMSE. Thus the biased model is
displaced mainly by ME*, whereas the noisy model is higher on SDE*.

```{r solar}
gg_solar(models, obs, label = TRUE, x.axis_begin = -1.5,
         x.axis_end = 1.5, y.axis_end = 2, by = 0.25)
```

## Target diagram

The target diagram uses the same error decomposition but gives SDE a sign, so
it distinguishes predictions that are smoother than observations from those
that are too variable. Distance from the origin remains standardized RMSE.

```{r target}
gg_target(models, obs, label = TRUE, axis_begin = -2, axis_end = 2, by = 0.5)
```

## Customisation

```{r customise}
gg_taylor(models, obs) +
  ggplot2::labs(title = "Model comparison") +
  ggplot2::theme_minimal()
```

Wadoux, A. M. J.-C., Walvoort, D. J. J., and Brus, D. J. (2022). An
integrated approach for the evaluation of quantitative soil maps through
Taylor and solar diagrams. *Geoderma*, 405, 115332.
<https://doi.org/10.1016/j.geoderma.2021.115332>
